---
title: "Population density of Münster with a small regression model to predict the future"
output: html_notebook
---

Easy regression problem for the population density in Münster, Germany. The goal is to load data from a publicly accessable pdf file and predict the future population of the city.

# Installation
```{r}
# install our package manager to online all our dependencies
if (!require("pacman")) install.packages("pacman"); library(pacman)
pacman::p_load(pdftools, stringr, ggplot2, reshape2, mice)
```

# Load Data

Next we load the PDF and extract the lines of importance to us.

```{r}
pdfTablePageForAllYears <- pdf_text("https://www.stadt-muenster.de/fileadmin//user_upload/stadt-muenster/61_stadtentwicklung/pdf/jahr/Jahres-Statistik_2016_Bevoelkerung.pdf","./71_PV.62.pdf")[14]
sampleLineSplit <- strsplit(pdfTablePageForAllYears, "\r\n")
# Format is in a list, so we first access list index and lines afterwards
filterRelevantLines <- sampleLineSplit[[1]][8:52]
head(filterRelevantLines, n = 2)
tail(filterRelevantLines, n = 2)
```

# Preprocessing

We have a small problem with numbers, since there is a whitespace in the population number, lets clean that up first. Afterwards we create clean strings in a vector of year and populationnumber.

```{r}
populationToNumber <- str_replace_all(filterRelevantLines, "(\\d{2,3}) (\\d{3})", "\\1\\2")
head(populationToNumber, n = 4)
# Match 4 Digits with an optional * for merged years, 0 to x whitespaces and then either a . or a population number.
yearAndPopulationExtract <- unlist(str_extract_all(populationToNumber,"\\d{4}\\*?\\s+(\\d{5,6})?\\.?"))
head(yearAndPopulationExtract, n = 4)
```

Now, on to create a dataframe, this means preprocessing our data even further into year|population, since we have a bonus condition "Eingemeindungen" (merging of communities) we want to use that as well, which is why we create one more column "merged".

```{r}
years = as.integer(c())
population = as.integer(c())
merged = c()
# Write data into empty vectors
for(data in yearAndPopulationExtract){
  if (grepl("\\*", data)){
    merged <- c(merged,T)
  } else {
    merged <- c(merged,F)
  }
  splitYearsAndPop <- unlist(strsplit(data,"\\*?\\s+"))
  years <- c(years,splitYearsAndPop[1])
  population <- c(population,splitYearsAndPop[2])
}
stopifnot(length(years) == length(merged) && length(merged) == length(population))
allDataInFrame <- data.frame(years,population,merged)
allDataInFrame <- allDataInFrame[order(years),] 
# check the table
knitr::kable(allDataInFrame[1:5,], caption = "A Knitr table.", floating.environment="sidewaystable" )
```

We now have a few characters of "." in our data, lets make it so that they are NA.

```{r}
allDataInFrame[allDataInFrame=="."]<-NA
knitr::kable(allDataInFrame[1:5,], caption = "A Knitr table.", floating.environment="sidewaystable" )
```

Now, we train a linear model to impute the missing values. We remove the last line since we want to predict the future later on.

```{r}
sapply(allDataInFrame, class)
# years and population are of type factor, which is why we use this workaround
allDataInFrame$years <- as.integer(as.character(allDataInFrame$years))
allDataInFrame$population <- as.integer(as.character(allDataInFrame$population))
allDataInFrameWithoutLastRow <- allDataInFrame[-nrow(allDataInFrame),]
# impute using mice, data looks weird for 1873, but we'll take it...
imp <- mice(allDataInFrameWithoutLastRow, method = "pmm", m = 1, maxit = 50)
allDataInFrameWithoutLastRow <- complete(imp)
allDataInFrameWithoutLastRow$years <- as.integer(as.character(allDataInFrameWithoutLastRow$years))
allDataInFrameWithoutLastRow$population <- as.integer(as.character(allDataInFrameWithoutLastRow$population))
```

# Tests

# Visualization

Now lets plot, whatever we did here.

```{r}
scatter.smooth(x=allDataInFrameWithoutLastRow$years, y=allDataInFrameWithoutLastRow$population, main="Münster Population Over Time") 
```

Quite some outliers, interesting to see, how much damage second world war did, while the first world war didn't seem to have any significance.

We won't do any evaluation on the model we're building, let's just see what happens, if we want to predict years 2016-2025.


```{r}
predictionFrame <- data.frame("years" = 1812:2012, "population" = 0)
lmMod <- lm(population ~ poly(years,3), data=allDataInFrameWithoutLastRow)  # build the model
predictionFrame$population = predict(lmMod, predictionFrame) 
plot(allDataInFrameWithoutLastRow$years,allDataInFrameWithoutLastRow$population,type='l',col='navy',main='Nonlinear relationship',lwd=5)
lines(predictionFrame$years, predictionFrame$population,col='red',lwd=5)
```

Not too bad, looks alright for now, we have to test some parameters like the AUC to optimize the model.



